---
title: "Emission Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Emission Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mdgm)
library(ggplot2)
```

## Overview

In a hierarchical MDGM, the spatial field $z$ is latent and observations $y$
are generated through an **emission distribution**. The mdgm package supports
three emission families:

| Family | Observation type | Parameters | Prior |
|:-------|:-----------------|:-----------|:------|
| Bernoulli | Binary (0/1) | $p_k$ (success probability) | Beta$(a, b)$ |
| Gaussian | Continuous | $\mu_k$, $\sigma_k^2$ | Independent Normal, Inverse-Gamma |
| Poisson | Count | $\lambda_k$ (rate) | Gamma$(\alpha, \beta)$ |

All emission families enforce an **identifiability constraint** on their
location parameters: $p_0 < p_1 < \cdots$ (Bernoulli),
$\mu_0 < \mu_1 < \cdots$ (Gaussian), or
$\lambda_0 < \lambda_1 < \cdots$ (Poisson). This is achieved via truncated
conjugate posterior sampling.

## Simulating a spatial field

For the examples below we use a simple region-growing algorithm to generate
irregular spatial clusters on a grid graph. Starting from random seed vertices,
regions expand outward one cell at a time, producing blocky but asymmetric
patterns:

```{r region-grow}
grow_regions <- function(nug, n_colors, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  n <- nug$nvertices()
  z <- rep(NA_integer_, n)

  # Plant one seed per color
  seeds <- sample(n, n_colors)
  for (k in seq_along(seeds)) z[seeds[k]] <- k - 1L

  # Grow: repeatedly claim a random unclaimed neighbor
  repeat {
    boundary <- which(!is.na(z))
    candidates <- integer(0)
    for (v in boundary) {
      nbrs <- nug$neighbors(v)
      candidates <- c(candidates, nbrs[is.na(z[nbrs])])
    }
    candidates <- unique(candidates)
    if (length(candidates) == 0) break

    pick <- sample(candidates, 1)
    claimed <- nug$neighbors(pick)
    claimed <- claimed[!is.na(z[claimed])]
    z[pick] <- z[sample(claimed, 1)]
  }
  z
}
```

```{r sim-field, fig.width = 4, fig.height = 4}
nug <- nug_from_grid(8, 8, seed = 42L)
n <- nug$nvertices()

z_true <- grow_regions(nug, n_colors = 2, seed = 7)

grid_df <- data.frame(
  x = rep(1:8, 8), y = rep(8:1, each = 8),
  z = factor(z_true)
)
ggplot(grid_df, aes(x, y, fill = z)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_manual(values = c("0" = "#440154", "1" = "#fde725")) +
  coord_equal() + theme_minimal() +
  labs(title = "Simulated spatial field (region growing)", fill = "z")
```

## Bernoulli emission

The Bernoulli emission models binary observations. Each vertex $i$ can have
multiple replicate observations $y_{i1}, \ldots, y_{im_i}$, each drawn
independently from $\text{Bernoulli}(p_{z_i})$. Multiple replicates per vertex
are needed for identifiability in the Bernoulli case.

### Example: binary indicators on a grid

```{r bernoulli-sim}
set.seed(1)
p_true <- c(0.3, 0.7)

# 5 replicate observations per vertex
y_bern <- lapply(seq_len(n), function(i) {
  rbinom(5, 1, p_true[z_true[i] + 1])
})
```

Fit the model:

```{r bernoulli-fit}
model_b <- mdgm_model(nug, dag_type = "spanning_tree",
                       n_colors = 2L, emission = "bernoulli")

result_b <- mcmc(model_b, y = y_bern,
                 z_init = sample(0:1, n, replace = TRUE),
                 psi_init = 0.5,
                 theta_init = c(0.3, 0.7),
                 n_iter = 5000L,
                 psi_tune = 1.0,
                 seed = 42L,
                 nug = nug)

result_b$summary(burnin = 1000L)
```

### Posterior trace plots

```{r bernoulli-trace, fig.width = 6, fig.height = 4}
ep_b <- result_b$emission_params()
n_iter <- ncol(ep_b$p)

p_df <- data.frame(
  iteration = rep(seq_len(n_iter), 2),
  value = c(ep_b$p[1, ], ep_b$p[2, ]),
  parameter = rep(c("p_1", "p_2"), each = n_iter)
)

ggplot(p_df, aes(iteration, value, color = parameter)) +
  geom_line(alpha = 0.5, linewidth = 0.3) +
  geom_hline(yintercept = p_true, linetype = "dashed", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Bernoulli emission: p trace",
       x = "Iteration", y = "Value", color = NULL)
```

### Edge inclusion probabilities

```{r bernoulli-eip, fig.width = 5, fig.height = 5, eval = requireNamespace("igraph", quietly = TRUE)}
result_b$plot(burnin = 1000L, which = "edge_inclusion")
```

## Gaussian emission

The Gaussian emission models continuous observations. Each observation is drawn
from $\mathcal{N}(\mu_{z_i}, \sigma_{z_i}^2)$. A single observation per vertex
is sufficient for identifiability.

The priors on $\mu_k$ and $\sigma_k^2$ are independent:
$$\mu_k \sim \mathcal{N}(\mu_0, \sigma^2_0), \qquad \sigma_k^2 \sim \text{InverseGamma}(\alpha_0, \beta_0)$$
where $\sigma^2_0$ is the prior variance for the mean.

### Example: spatial temperature field

Here we use overlapping emission distributions — the two groups differ in mean
but share similar spread, making classification rely on both the emission
signal and the spatial structure:

```{r gaussian-sim, fig.width = 5, fig.height = 4}
set.seed(2)
mu_true <- c(5, 12)
sigma2_true <- c(9, 9)

# Single observation per vertex
y_gauss <- lapply(seq_len(n), function(i) {
  rnorm(1, mu_true[z_true[i] + 1], sqrt(sigma2_true[z_true[i] + 1]))
})

# Visualize the observations
obs_df <- data.frame(
  x = rep(1:8, 8), y = rep(8:1, each = 8),
  value = vapply(y_gauss, `[`, numeric(1), 1)
)
ggplot(obs_df, aes(x, y, fill = value)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient2(low = "#440154", mid = "#21918c", high = "#fde725",
                       midpoint = mean(obs_df$value)) +
  coord_equal() + theme_minimal() +
  labs(title = "Observed temperatures (single obs per vertex)", fill = "y")
```

Fit with Gaussian emission:

```{r gaussian-fit}
model_g <- mdgm_model(nug, dag_type = "spanning_tree",
                       n_colors = 2L, emission = "gaussian")

# theta_init: c(mu_1, mu_2, sigma2_1, sigma2_2)
result_g <- mcmc(model_g, y = y_gauss,
                 z_init = sample(0:1, n, replace = TRUE),
                 psi_init = 0.5,
                 theta_init = c(4, 14, 9, 9),
                 n_iter = 5000L,
                 psi_tune = 1.0,
                 seed = 42L,
                 nug = nug)

result_g$summary(burnin = 1000L)
```

### Posterior emission parameters

```{r gaussian-trace, fig.width = 6, fig.height = 4}
ep_g <- result_g$emission_params()
n_iter_g <- ncol(ep_g$mu)

mu_df <- data.frame(
  iteration = rep(seq_len(n_iter_g), 2),
  value = c(ep_g$mu[1, ], ep_g$mu[2, ]),
  parameter = rep(c("mu_1", "mu_2"), each = n_iter_g)
)

ggplot(mu_df, aes(iteration, value, color = parameter)) +
  geom_line(alpha = 0.5, linewidth = 0.3) +
  geom_hline(yintercept = mu_true, linetype = "dashed", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Gaussian emission: mu trace",
       x = "Iteration", y = "Value", color = NULL)
```

```{r gaussian-sigma2-trace, fig.width = 6, fig.height = 4}
sigma2_df <- data.frame(
  iteration = rep(seq_len(n_iter_g), 2),
  value = c(ep_g$sigma2[1, ], ep_g$sigma2[2, ]),
  parameter = rep(c("sigma2_1", "sigma2_2"), each = n_iter_g)
)

ggplot(sigma2_df, aes(iteration, value, color = parameter)) +
  geom_line(alpha = 0.5, linewidth = 0.3) +
  geom_hline(yintercept = sigma2_true, linetype = "dashed", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Gaussian emission: sigma2 trace",
       x = "Iteration", y = "Value", color = NULL)
```

### Edge inclusion probabilities

```{r gaussian-eip, fig.width = 5, fig.height = 5, eval = requireNamespace("igraph", quietly = TRUE)}
result_g$plot(burnin = 1000L, which = "edge_inclusion")
```

## Poisson emission

The Poisson emission models count data. Each observation is drawn from
$\text{Poisson}(\lambda_{z_i})$. As with the Gaussian case, a single
observation per vertex is sufficient for identifiability.

The prior is a Gamma distribution: $\lambda_k \sim \text{Gamma}(\alpha_0, \beta_0)$
with rate parameterization (mean $= \alpha_0 / \beta_0$).

### Example: spatial species counts

We again use overlapping rates so that the spatial structure contributes
meaningfully to classification:

```{r poisson-sim, fig.width = 5, fig.height = 4}
set.seed(3)
lambda_true <- c(4, 10)

# Single count per vertex
y_pois <- lapply(seq_len(n), function(i) {
  rpois(1, lambda_true[z_true[i] + 1])
})

obs_df_p <- data.frame(
  x = rep(1:8, 8), y = rep(8:1, each = 8),
  value = vapply(y_pois, `[`, numeric(1), 1)
)
ggplot(obs_df_p, aes(x, y, fill = value)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient(low = "#440154", high = "#fde725") +
  coord_equal() + theme_minimal() +
  labs(title = "Observed counts (single obs per vertex)", fill = "y")
```

Fit with Poisson emission:

```{r poisson-fit}
model_p <- mdgm_model(nug, dag_type = "spanning_tree",
                       n_colors = 2L, emission = "poisson")

result_p <- mcmc(model_p, y = y_pois,
                 z_init = sample(0:1, n, replace = TRUE),
                 psi_init = 0.5,
                 theta_init = c(3, 8),
                 n_iter = 5000L,
                 psi_tune = 1.0,
                 seed = 42L,
                 nug = nug)

result_p$summary(burnin = 1000L)
```

### Posterior trace plots

```{r poisson-trace, fig.width = 6, fig.height = 4}
ep_p <- result_p$emission_params()
n_iter_p <- ncol(ep_p$lambda)

lambda_df <- data.frame(
  iteration = rep(seq_len(n_iter_p), 2),
  value = c(ep_p$lambda[1, ], ep_p$lambda[2, ]),
  parameter = rep(c("lambda_1", "lambda_2"), each = n_iter_p)
)

ggplot(lambda_df, aes(iteration, value, color = parameter)) +
  geom_line(alpha = 0.5, linewidth = 0.3) +
  geom_hline(yintercept = lambda_true, linetype = "dashed", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Poisson emission: lambda trace",
       x = "Iteration", y = "Value", color = NULL)
```

### Edge inclusion probabilities

```{r poisson-eip, fig.width = 5, fig.height = 5, eval = requireNamespace("igraph", quietly = TRUE)}
result_p$plot(burnin = 1000L, which = "edge_inclusion")
```

## Comparing latent field recovery

For all three emission types, the posterior mode of the latent field should
recover the true spatial pattern:

```{r field-recovery, fig.width = 7, fig.height = 3.5}
burnin <- 1000L

recover_field <- function(result, burnin) {
  z_post <- result$z()
  z_burn <- z_post[, (burnin + 1):ncol(z_post)]
  apply(z_burn, 1, function(row) {
    tbl <- tabulate(row + 1L, nbins = 2)
    which.max(tbl) - 1L
  })
}

z_mode_b <- recover_field(result_b, burnin)
z_mode_g <- recover_field(result_g, burnin)
z_mode_p <- recover_field(result_p, burnin)

field_df <- data.frame(
  x = rep(rep(1:8, 8), 4),
  y = rep(rep(8:1, each = 8), 4),
  value = c(z_true, z_mode_b, z_mode_g, z_mode_p),
  panel = rep(c("True field", "Bernoulli", "Gaussian", "Poisson"),
              each = n)
)
field_df$panel <- factor(field_df$panel,
                         levels = c("True field", "Bernoulli",
                                    "Gaussian", "Poisson"))

ggplot(field_df, aes(x, y, fill = factor(value))) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_manual(values = c("0" = "#440154", "1" = "#fde725")) +
  coord_equal() +
  facet_wrap(~panel, nrow = 1) +
  theme_minimal() +
  labs(fill = "z")
```

## Choosing an emission family

| Scenario | Recommended emission |
|:---------|:--------------------|
| Binary outcomes (presence/absence, success/failure) | `"bernoulli"` |
| Continuous measurements (temperature, concentration) | `"gaussian"` |
| Count data (species counts, event counts) | `"poisson"` |

## Prior tuning tips

- **Bernoulli** `c(a, b)`: Use `c(1, 1)` (uniform) as a default. Increase
  `a` and `b` to shrink $p_k$ toward 0.5 if you expect weak signal.
- **Gaussian** `c(mu_0, sigma2_0, alpha_0, beta_0)`: Set `mu_0` near the data
  mean, `sigma2_0` large (e.g., 10000) for a vague location prior, and
  `alpha_0 = 2`, `beta_0` near the expected variance for a weakly informative
  scale prior.
- **Poisson** `c(alpha_0, beta_0)`: The prior mean is `alpha_0 / beta_0`.
  Use `c(1, 0.1)` for a diffuse prior with mean 10, or match to the expected
  count range.
