---
title: "Getting Started with mdgm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mdgm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mdgm)
```

## What is mdgm?

The **mdgm** package implements Bayesian inference for discrete spatial random
fields using **Mixtures of Directed Graphical Models**. Like a Markov random
field (MRF), the MDGM takes an undirected graph as input to represent spatial
structure. However, instead of specifying a single joint distribution with an
intractable normalizing constant, the MDGM defines a mixture over directed
acyclic graphs (DAGs) compatible with the undirected graph. Each DAG admits an
efficient factorization, avoiding the computational burden of the MRF partition
function while providing valid probabilistic inference.

The approach is described in:

> Carter, J. B. and Calder, C. A. (2024). Mixture of Directed Graphical Models
> for Discrete Spatial Random Fields.
> [arXiv:2406.15700](https://arxiv.org/abs/2406.15700)

## Installation

Install the development version from GitHub:

```r
# install.packages("remotes")
remotes::install_github("jbcart/mdgm")
```

The package requires a C++20 compiler.

## Quick example

The basic workflow is:

1. **Build a graph** — Define the undirected neighborhood structure.
2. **Specify a model** — Choose the DAG type and (optionally) an emission
   distribution.
3. **Run MCMC** — Sample from the posterior.
4. **Analyze results** — Inspect posterior summaries, edge inclusion
   probabilities, etc.

### Step 1: Build a graph

Create a 4x4 grid graph using an adjacency matrix with rook neighbors:

```{r grid-graph}
n_row <- 4
n_col <- 4
n <- n_row * n_col
A <- matrix(0, n, n)
for (i in 1:n) {
  ri <- (i - 1) %/% n_col + 1
  ci <- (i - 1) %% n_col + 1
  for (j in 1:n) {
    rj <- (j - 1) %/% n_col + 1
    cj <- (j - 1) %% n_col + 1
    if (abs(ri - rj) + abs(ci - cj) == 1) {
      A[i, j] <- 1
    }
  }
}
nug <- nug_from_adj_mat(A, seed = 42L)
nug$nvertices()
nug$nedges()
```

### Step 2: Define a model

Create a hierarchical model with spanning-tree DAGs and a Bernoulli emission:

```{r model}
model <- mdgm_model(nug, dag_type = "spanning_tree", n_colors = 2L,
                     emission = "bernoulli")
model$has_emission()
model$ncolors()
```

For a **standalone** model (where the spatial field is observed directly), omit
the `emission` argument:

```r
model_standalone <- mdgm_model(nug, dag_type = "spanning_tree")
```

### Step 3: Run MCMC

Simulate some binary observations and run the sampler:

```{r mcmc}
set.seed(1)
# Simulate observations: vertex i gets a few Bernoulli draws
y <- lapply(1:n, function(i) rbinom(5, 1, ifelse(i <= 8, 0.2, 0.8)))

# Initial values
z_init <- rep(0:1, each = 8)
result <- mcmc(model, y = y, z_init = z_init, psi_init = 0.5,
               eta_init = c(0.3, 0.7), n_iter = 500L, seed = 42L)
```

### Step 4: Analyze results

```{r results}
result$summary()
```

Access posterior samples directly:

```{r samples}
# Psi trace
psi_samples <- result$psi()
plot(psi_samples, type = "l", xlab = "Iteration", ylab = expression(psi),
     main = "Psi trace plot")

# Acceptance rates
result$acceptance_rates()
```

Compute edge inclusion probabilities (discarding the first 100 iterations as
burn-in):

```{r edge-probs}
eip <- result$edge_inclusion_probs(nug, burnin = 100L)
head(eip[order(-eip$prob), ])
```

See the [MDGM Model Specification](mdgm-models.html) vignette for details on
the statistical model and DAG types.
