---
title: "Getting Started with mdgm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mdgm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mdgm)
library(ggplot2)
```

## What is mdgm?

The **mdgm** package implements Bayesian inference for discrete spatial random
fields using **Mixtures of Directed Graphical Models**. Like a Markov random
field (MRF), the MDGM takes an undirected graph as input to represent spatial
structure. However, instead of specifying a single joint distribution with an
intractable normalizing constant, the MDGM defines a mixture over directed
acyclic graphs (DAGs) compatible with the undirected graph. Each DAG admits an
efficient factorization, avoiding the computational burden of the MRF partition
function while providing valid probabilistic inference.

The approach is described in:

> Carter, J. B. and Calder, C. A. (2024). Mixture of Directed Graphical Models
> for Discrete Spatial Random Fields.
> [arXiv:2406.15700](https://arxiv.org/abs/2406.15700)

## Building a graph

Create a 4x4 grid graph with rook adjacency using `nug_from_grid()`:

```{r grid-graph}
nug <- nug_from_grid(4, 4, seed = 42L)
nug$nvertices()
nug$nedges()
```

Graphs can also be constructed from an adjacency matrix, edge list, or
adjacency list — see the
[Working with Undirected Graphs](graph-construction.html) vignette.

## Fitting a standalone model

In a standalone model, the spatial field $z$ is observed directly — no emission
distribution is needed. Here we fit a spanning-tree MDGM to a deterministic
checkerboard pattern on the 4x4 grid:

```{r standalone-fit}
z <- c(1L, 1L, 0L, 0L,
       1L, 1L, 0L, 0L,
       0L, 0L, 1L, 1L,
       0L, 0L, 1L, 1L)

model <- mdgm_model(nug, dag_type = "spanning_tree")

result <- mcmc(model, z_init = z, psi_init = 0.5,
               n_iter = 2000L, psi_tune = 1.0, seed = 42L)
result$summary()
```

Visualize the spatial field as a raster:

```{r grid-plot, fig.width = 4, fig.height = 4}
grid_df <- data.frame(
  x = rep(1:4, times = 4),
  y = rep(4:1, each = 4),
  z = factor(z)
)

ggplot(grid_df, aes(x, y, fill = z)) +
  geom_raster() +
  scale_fill_manual(values = c("0" = "#440154", "1" = "#fde725")) +
  coord_equal() +
  theme_minimal() +
  labs(title = "4x4 Spatial Field", fill = "z")
```

## Posterior diagnostics

Trace plot for the spatial dependence parameter $\psi$:

```{r psi-trace, fig.width = 6, fig.height = 3}
psi_df <- data.frame(iteration = seq_along(result$psi()), psi = result$psi())

ggplot(psi_df, aes(iteration, psi)) +
  geom_line(linewidth = 0.3) +
  theme_minimal() +
  labs(x = "Iteration", y = expression(psi), title = "Psi trace plot")
```

Acceptance rates:

```{r acceptance}
result$acceptance_rates()
```

## Visualizing graph structure with igraph

The [igraph](https://r.igraph.org/) package (suggested, not required) can plot
the neighborhood graph:

```{r igraph-grid, fig.width = 4, fig.height = 4, eval = requireNamespace("igraph", quietly = TRUE)}
library(igraph)

# Build igraph object from the NUG edge structure
n <- nug$nvertices()
el <- do.call(rbind, lapply(1:n, function(v) {
  nbrs <- nug$neighbors(v)
  nbrs <- nbrs[nbrs > v]
  if (length(nbrs) == 0) return(NULL)
  cbind(v, nbrs)
}))
g <- graph_from_edgelist(el, directed = FALSE)

# Grid layout
coords <- cbind((seq_len(n) - 1) %% 4 + 1, 4 - (seq_len(n) - 1) %/% 4)

plot(g, layout = coords, vertex.size = 20, vertex.label = 1:n,
     vertex.color = c("#440154", "#fde725")[z + 1],
     vertex.label.color = "white", edge.color = "grey40",
     main = "4x4 Grid (NUG)")
```

## Edge inclusion probabilities

The `edge_inclusion_probs()` method counts how often each undirected edge
appears in the posterior spanning-tree samples. We can use these proportions to
scale edge widths:

```{r edge-inclusion, fig.width = 4, fig.height = 4, eval = requireNamespace("igraph", quietly = TRUE)}
eip <- result$edge_inclusion_probs(nug, burnin = 200L)
head(eip[order(-eip$prob), ])

# Match edge inclusion probs to igraph edge ordering
edge_probs <- numeric(ecount(g))
for (i in seq_len(nrow(eip))) {
  eid <- get.edge.ids(g, c(eip$vertex1[i], eip$vertex2[i]))
  edge_probs[eid] <- eip$prob[i]
}

plot(g, layout = coords, vertex.size = 20, vertex.label = 1:n,
     vertex.color = c("#440154", "#fde725")[z + 1],
     vertex.label.color = "white",
     edge.width = edge_probs * 10,
     edge.color = "grey30",
     main = "Edge inclusion probabilities")
```

Edges between same-colored vertices appear more frequently in the posterior
spanning trees, reflecting the spatial dependence captured by $\psi$.

## Fitting a hierarchical model

In a hierarchical model, $z$ is a latent field and observations are generated
through an emission distribution. Here we use a Bernoulli emission:

```{r hierarchical-fit}
model_h <- mdgm_model(nug, dag_type = "spanning_tree", n_colors = 2L,
                       emission = "bernoulli")

set.seed(1)
y <- lapply(1:n, function(i) rbinom(5, 1, ifelse(i <= 8, 0.2, 0.8)))

result_h <- mcmc(model_h, y = y, z_init = rep(0:1, each = 8), psi_init = 0.5,
                 eta_init = c(0.3, 0.7), n_iter = 500L, seed = 42L)
result_h$summary()
```

## Next steps

- **[Working with Undirected Graphs](graph-construction.html)** — Graph
  constructors, querying structure, and sampling spanning trees.
- **[MDGM Model Specification](mdgm-models.html)** — DAG types, the spatial
  field model, emission distributions, and MCMC details.
